### 2022-09-22 AISTATS2023/paper.tex
added content in Synthetic Examples

\subsection{Examples 2: Bin-Splitting}
\label{sec:simulation-examples-2}

The aim of this example is to illustrate the advantages of automatic
bin-splitting based on the objective of Eq.(\ref{eq:opt}). For
evaluating our approach, we create the following validation framework;
We generate a big dataset using dense sampling \(N=10^6\) and we treat
as ground-truth the ALE estimation using dense fixed-size binning
\(K=10^3\). Afterwards, we generate less samples \(X=500\) and we
compare (a) the fixed-size ALE estimation for many different \(K\)
versus (b) the auto-bin splitting algorithm. In all cases, the data
generating distribution is \(p(\mathbf{x}) = p(x_2|x_1)p(x_1)\) where
\(x_1 \sim \mathcal{U}(0,1)\),
\(x_2 \sim \mathcal{N}(x_1, \sigma_2^2=2)\).

We denote as
\(\mathcal{Z^{\mathtt{DP}}} = \{z^{\mathtt{DP}}_{k-1}, \cdots,
z^{\mathtt{DP}}_{K}\}\) the sequence obtained by automatic-splitting
based on the optimisation problem of Eq.~(\ref{eq:opt}) and with
\(\mathcal{Z^{\mathtt{K}}}\}\) the corresponding fixed-sized
splitting. The comparison is done using three metrics; (a)
\(\mathcal{L}_{\mathtt{DP|K}}^{\mu} =
\frac{1}{|\mathcal{Z}^{\mathtt{DP|K}}|} \sum_{k=1}^K | \mu_k -
\hat{\mu}_k | \) quantifies the average error in estimating the the
expected effect, (b)
\(\mathcal{L}_{\mathtt{DP|K}}^{\sigma^2} =
\frac{1}{|\mathcal{Z}^{\mathtt{DP|K}}|} \sum_{k=1}^K | \sigma_k^2 -
\hat{\sigma}^2_k | \) the average error in estimating the variance of
the effect and (c)
\(\mathcal{L}^{\rho^2}_{\mathtt{DP|K}} =
\frac{1}{|\mathcal{Z}^{\mathtt{DP|K}}|} \sum_{k=1}^K \rho^2_k \) the
average extra nuisance uncertainty.

\paragraph{Example 1.}

In this example, we defin a black-box function
\(f_1(x_1) = a_1x_1 + x_1x_2\) with 6 piecewise-linear regions of
different-size, i.e., \(a_1\) equals to \(=\{2, -2, 5, -10, 0.5\}\) in
the regions defined by the sequence \(\{0, 0.2, 0.4, 0.45, 0.5,
1\}\). The example highlights the importance of variable-size
splitting. Given that the smalest intervals size is \(0.05\) any
fixed-size splitting with less than \( \frac{1}{\Delta x} = 20\) bins,
will mix together effects and increase the variance error
\(\mathcal{L}_{\mathtt{K}}^{\sigma^2}\) due to the increase of the
residual \(\mathcal{L}_{\mathtt{K}}^{\rho^2} > 0\). This is evident in
figure. On the other hand, a large number of bins divides homogeneuos
regions, i.e. region \([0.5, 1)\), into smaller ones with more
unstable estimations which increases
\(\mathcal{L}_{\mathtt{K}}^{\mu}\) Figure. Variable-size binning leads
to the optimal solution. As we see in figure automatic-bin splitting
small bins in areas with fine-grain effects and merges together
regions with similar effects, leading to a near optimal bin splitting.

\begin{figure}[h]
  \centering
  \includegraphics[width=.23\textwidth]{example_bin_splitting_1/fig_1.pdf}
  \includegraphics[width=.23\textwidth]{example_bin_splitting_1/fig_2.pdf}\\
  \includegraphics[width=.23\textwidth]{example_bin_splitting_1/fig_3.pdf}
  \includegraphics[width=.23\textwidth]{example_bin_splitting_1/fig_4.pdf}
  \caption{Figure 1}
  \label{fig:ex-synth-1-1}
\end{figure}

\paragraph{Example 2.}

The black-box function \(f(\mathbf{x}) = 4x_1^2 + x_2^2 + x_1x_2\)

\begin{figure}[h]
  \centering
  \includegraphics[width=.23\textwidth]{example_bin_splitting_2/fig_1.pdf}
  \includegraphics[width=.23\textwidth]{example_bin_splitting_2/fig_2.pdf}\\
  \includegraphics[width=.23\textwidth]{example_bin_splitting_2/fig_3.pdf}
  \includegraphics[width=.23\textwidth]{example_bin_splitting_2/fig_4.pdf}
  \caption{Figure 1}
  \label{fig:ex-synth-1-1}
\end{figure}

@@

>>>>

\subsection{Case 2: Bin-Splitting}
\label{sec:simulation-examples-2}

In this simulation, we aim to to quantify the advantages of automatic
bin-splitting based on the objective of Eq.(\ref{eq:opt}), through the
following validation framework. We generate a big dataset with dense
sampling (\(N=10^6\)) and we treat the DALE estimation with dense
fixed-size bins (\(K=10^3\)) as ground-truth. Afterwards, we generate
less samples (\(N=500\)) and we compare (a) the fixed-size DALE
estimation for many different \(K\) versus (b) the auto-bin splitting
algorithm. In all cases, we use a bivariate black-box function
\(f(\cdot)\) where the samples are instances of the distribution
\(p(\mathbf{x}) = p(x_2|x_1)p(x_1)\) where
\(x_1 \sim \mathcal{U}(0,1)\) and
\(x_2 \sim \mathcal{N}(x_1, \sigma_2^2=0.5)\).

We denote as
\(\mathcal{Z^{\mathtt{DP}}} = \{z^{\mathtt{DP}}_{k-1}, \cdots,
z^{\mathtt{DP}}_{K}\}\) the sequence obtained by automatic
bin-splitting based on the optimisation problem of Eq.~(\ref{eq:opt})
and with \(\mathcal{Z^{\mathtt{K}}}\) the fixed-size splitting with
\(K\) bins. The evaluation is done with regard to two metrics;
\(\mathcal{L}_{\mathtt{DP|K}}^{\mu} =
\frac{1}{|\mathcal{Z}^{\mathtt{DP|K}}|} \sum_{k \in
  \mathcal{Z}^{\mathtt{DP|K}}} | \mu_k - \hat{\mu}_k | \) quantifies
the average error in estimating the the expected effect and
\(\mathcal{L}_{\mathtt{DP|K}}^{\sigma^2} =
\frac{1}{|\mathcal{Z}^{\mathtt{DP|K}}|} \sum_{k \in
  \mathcal{Z}^{\mathtt{DP|K}}} | \sigma_k^2 - \hat{\sigma}^2_k | \)
the average error in estimating the variance of the effect. The metric
\(\mathcal{L}^{\rho^2}_{\mathtt{DP|K}} =
\frac{1}{|\mathcal{Z}^{\mathtt{DP|K}}|} \sum_{k \in
  \mathcal{Z}^{\mathtt{DP|K}}} \rho^2_k \) quantifies the average
nuisance uncertainty. The optimal bin-splitting is the one that
minimises, at the same time, both \(\mathcal{L}^{\mu}\) and
\(\mathcal{L}^{\sigma^2}\) errors. For consistent results, in all the
examples below, we regenerate samples and repeat the computations for
\(t = 10\) times, providing the mean value for all metrics.

\paragraph{Piecewise-Linear Function.}

In this example, we define \(f(\mathbf{x}) = a_1x_1 + x_1x_2\) with
\(5\) piecewise-linear regions of different-size, i.e., \(a_1\) equals
to \(=\{2, -2, 5, -10, 0.5\}\) in the intervals defined by the
sequence \(\{0, 0.2, 0.4, 0.45, 0.5, 1\}\). As we observe, in
Figure~\ref{fig:ex-synth-2-1}, NAME method extracts a sequence of
intervals with better \(\mathcal{L}_{\mathtt{DP}}^{\mu}\) and
\(\mathcal{L}_{\mathtt{DP}}^{\sigma^2}\) error compared to any
fixed-size splitting. Analyzing the fixed-size errors helps us
understand the importance of variable-size splitting. In
Figure~\ref{fig:ex-synth-2-1}(b), we observe a positive trend between
\(\mathcal{L}^{\mu}_{\mathtt{K}}\) and \(K\), concluding that bin
effect estimation is more incosistent as \(\Delta x\) becomes smaller,
due to less points contributing to each bin. The interpretation of
variance error is slightly more complex. Given that the smallest
interval is \(\Delta x = 0.05 \Rightarrow K = 20\) and all intervals
are multiples of the smallest interval, any \(K\) that is not a
multiple of \(20\) adds nuisance uncertainty
\(\mathcal{L}^{\rho^2}_{\mathtt{K}}\) leading to a high variance error
\(\mathcal{L}^{\sigma^2}_{\mathtt{K}}\). In these cases, the variance
error reduces as \(K\) grows bigger because the length of the bins
that lie in the limit between two piecewise linear regions becomes
smaller. For \(K=\{20, 40, 60, 80, 100\}\) where
\(\mathcal{L}^{\rho^2}_{\mathtt{K}} \approx 0\), we conclude the same
as with with the mean effect error, i.e. the estimation becomes more
incosistent as \(K\) grows larger.

Variable-size extracts correctly the fine-grain bins, e.g., intervals
\([0.4, 0.45]\), \([0.45, 0.5]\), and acts as a merging mechanism to
create wider bins when the effect remains constant, e.g. interval
\([0.5, 1]\), leading to an optimal solution.

\begin{figure}[h]
  \centering
  \includegraphics[width=.23\textwidth]{example_bin_splitting_1/fig_1.pdf}
  \includegraphics[width=.23\textwidth]{example_bin_splitting_1/fig_2.pdf}\\
  \includegraphics[width=.23\textwidth]{example_bin_splitting_1/fig_3.pdf}
  \includegraphics[width=.23\textwidth]{example_bin_splitting_1/fig_4.pdf}
  \caption{Figure 1}
  \label{fig:ex-synth-2-1}
\end{figure}

\paragraph{Non-Linear Function.}

In this example, we define a black-box function
\(f(\mathbf{x}) = 4x_1^2 + x_2^2 + x_1x_2\), where the effect is
non-linear in all the range of \(x\). This case has two
specialties. First, there is no obvious advantage of variable-size
versus fixed-size splitting, and, second, there is not an apriori
optimal bin-size. Widening a bin will increase the resolution error
\(\mathcal{L}^{\rho^2}\) and narrowing will make less robust. In
Figure~\ref{fig:ex-synth-2-2}, we observe that automatic bin splitting
finds a solution that compromises the conflicting objectives, i.e., it
keeps as low as possible both the main effect
\(\mathcal{L}^{\mu}_{\mathtt{K}}\) and the variance error
\(\mathcal{L}^{\sigma^2}_{\mathtt{K}}\).

\begin{figure}[h]
  \centering
  \includegraphics[width=.23\textwidth]{example_bin_splitting_2/fig_1.pdf}
  \includegraphics[width=.23\textwidth]{example_bin_splitting_2/fig_2.pdf}\\
  \includegraphics[width=.23\textwidth]{example_bin_splitting_2/fig_3.pdf}
  \includegraphics[width=.23\textwidth]{example_bin_splitting_2/fig_4.pdf}
  \caption{Figure 1}
  \label{fig:ex-synth-2-2}
\end{figure}

@@

\section{CONCLUSION}
